---
title: "Homework 5"
author: "Minghe Wang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rsample)
library(caret)
library(tidymodels)
library(e1071) 
library(ISLR)
library(factoextra)
library(gridExtra)
library(corrplot)
library(RColorBrewer) 
library(gplots)
library(jpeg)
```


# 1

```{r loadNsplit}
auto <- read_csv("./auto.csv")
#head(auto)

str(auto)
auto$origin <- as.factor(auto$origin)
auto$cylinders <- as.factor(auto$cylinders)
auto$mpg_cat <- as.factor(auto$mpg_cat)
contrasts(auto$mpg_cat)
auto <- na.omit(auto)

set.seed(1)
split_auto <- initial_split(auto, prop = 0.7)
training_data <- training(split_auto) 
testing_data <- testing(split_auto)
```

## a

```{r}
set.seed(1)
linear.tune <- tune.svm(mpg_cat ~ . , 
                        data = training_data, 
                        kernel = "linear", 
                        cost = exp(seq(-5,2, len = 50)),
                        scale = TRUE)

plot(linear.tune) # tuning curve
# summary(linear.tune)
linear.tune$best.parameters

best.linear <- linear.tune$best.model
summary(best.linear)

pred.train <- predict(best.linear, newdata = training_data)
pred.test <- predict(best.linear, newdata = testing_data)

cm.train <- confusionMatrix(data = pred.train, 
                reference = training_data$mpg_cat)
cm.test <- confusionMatrix(data = pred.test, 
                reference = testing_data$mpg_cat)

train_error <- 1 - cm.train$overall['Accuracy']
test_error <- 1 - cm.test$overall['Accuracy']
```

For linear SVM, train and test error rates are: `r train_error`, `r test_error`.

## b

```{r}
radial.tune <- tune.svm(mpg_cat ~ . , 
                        data = training_data, 
                        kernel = "radial", 
                        cost = exp(seq(1, 7, len = 50)),
                        gamma = exp(seq(-10, -2,len = 20)))

plot(radial.tune) # tuning curve
radial.tune$best.parameters

best.radial <- radial.tune$best.model
summary(best.radial)

pred.train_radial <- predict(best.radial, newdata = training_data)
pred.test_radial <- predict(best.radial, newdata = testing_data)

cm.train_radial <- confusionMatrix(data = pred.train_radial, 
                reference = training_data$mpg_cat)
cm.test_radial <- confusionMatrix(data = pred.test_radial, 
                reference = testing_data$mpg_cat)

train_error_radial <- 1 - cm.train_radial$overall['Accuracy']
test_error_radial <- 1 - cm.test_radial$overall['Accuracy']
```

For radial kernelized SVM, train and test error rates are: `r train_error_radial`, `r test_error_radial`.

# 2

## a

```{r}
data("USArrests")
#head(USArrests)
```

```{r}
hc.complete <- hclust(dist(USArrests), method = "complete")

fviz_dend(hc.complete, k = 3,        
          cex = 0.3, 
          palette = "jco", # color scheme; other palettes:"npg","aaas"...
          color_labels_by_k = TRUE,
          rect = TRUE, # whether to add a rectangle around groups.
          rect_fill = TRUE,
          rect_border = "jco",
          labels_track_height = 2.5) 

ind3.complete <- cutree(hc.complete, 3)
state1 <- rownames(USArrests[ind3.complete == 1,])
state1_names <- state1[!startsWith(state1, "NA")]

state2 <- rownames(USArrests[ind3.complete == 2,])
state2_names <- state2[!startsWith(state2, "NA")]

state3 <- rownames(USArrests[ind3.complete == 3,])
state3_names <- state3[!startsWith(state3, "NA")]
```

**Cluster 1 includes:** `r state1_names`

**Cluster 2 includes:** `r state2_names`

**Cluster 3 includes:** `r state3_names`

## b

```{r}
USArrests_scale <- scale(USArrests)

hc.complete_scaled <- hclust(dist(USArrests_scale), method = "complete")

fviz_dend(hc.complete_scaled, k = 3,        
          cex = 0.3, 
          palette = "jco", # color scheme; other palettes:"npg","aaas"...
          color_labels_by_k = TRUE,
          rect = TRUE, # whether to add a rectangle around groups.
          rect_fill = TRUE,
          rect_border = "jco",
          labels_track_height = 2.5) 

ind3.complete_scaled <- cutree(hc.complete_scaled, 3)
state1_scale <- rownames(USArrests_scale[ind3.complete_scaled == 1,])
state1_names_scale <- state1_scale[!startsWith(state1_scale, "NA")]

state2_scale <- rownames(USArrests_scale[ind3.complete_scaled == 2,])
state2_names_scale <- state2_scale[!startsWith(state2_scale, "NA")]

state3_scale <- rownames(USArrests_scale[ind3.complete_scaled == 3,])
state3_names_scale <- state3_scale[!startsWith(state3_scale, "NA")]
```

After scaling:

**Cluster 1 includes:** `r state1_names_scale`

**Cluster 2 includes:** `r state2_names_scale`

**Cluster 3 includes:** `r state3_names_scale`

Scaling the variables changes the clustering results significantly because without scaling, variables with larger absolute ranges will dominate the Euclidean distance calculations. Thus, clustering would primarily reflect variation in high-magnitude variables, not all variables equally.

Yes, they should be scaled — especially when variables are measured in different units or have different variances, which is the case in `USArrests`. This ensures that all variables contribute equally to the distance calculations, and clustering reflects balanced structure across all features.